Fast and Flexible Binary Instrumentation for the Kernel

### 2013-12-08 VEE-14/paper.tex
Hacking on the module wrappers section. Really chopping it down to the essentials.

@@

@@

\section{Scope-Driven Specialization}\label{sec:what}

Some instrumentation tools have a limited scope. That is, they are made to solve a specific problem within a specific domain. For example, Granary's Guardrail tool is concerned with finding bugs in Linux device drivers \cite{Guardrail}. Guardrail records and analyzes traces of device driver code to find bugs. However, Guardrail's analysis can be heavyweight, and so limiting the analysis to only device driver code results in better performance. Granary allows Guardrail to set the granularity of instrumentation to only instrument a specific module, while leaving the rest of the kernel to execute without overhead.

@@

Guardrail is an example of how tools can choose an instrumentation granularity that better matches their scope and needs. That is, a Granary tool can choose to instrument some code, while leaving other code to execute natively. This is useful when the instrumentation in question has a high overhead or when only a subset of the kernel's code is of interest. \Cref{tab:granularities} provides a high-level overview of the different granularities of instrumentation supported by Granary. Granary tools, however, are not restricted to only the Granularities listed in the table.

@@

\subsection{Instrumenting All Kernel Code}
Like DRK and btkernel, Granary can instrument the whole kernel. Instrumenting the whole kernel requires interposing on the system call entrypoint (x86-64 \texttt{MSR\_LSTAR} model-specific register) and on the interrupt vector entrypoints (stored in the interrupt descriptor table). For whole-kernel instrumentation, Granary boostraps by duplicating the the early entrypoint instructions that are responsible for switching execution onto a kernel stack, and then injects a jump to translate/instrumented kernel code after the stack switch. The translation process continues when a system call or interrupt occurs and follows the jump from duplicated into translated code. Granary only translate code that is known to operate on a kernel stack\footnote{Not all kernel code executes on a kernel stack. For example, when a user space application invokes the \texttt{syscall} instruction, control is transferred to the kernel, but no stack switch occurs. The kernel's early system call entrypoint code is responsible for switching execution onto a kernel stack.} so that instrumentation tools can safely save transient data in a reentrant way by spilling that data to the stack.

\begin{figure}[t]
\captionsetup[subfloat]{width=0.45\columnwidth}
\begin{tabularx}{\columnwidth}{XX}%
\subfloat[Full-kernel instrumentation.]{\hspace{2.5em}\epsfig{file=diagrams/kernel.eps,width=0.125\textwidth}} & %
\subfloat[Module-only instrumentation]{\hspace{2.5em}\epsfig{file=diagrams/module.eps,width=0.125\textwidth}} \\
\subfloat[Module-only instrumentation, with a function specialized to run natively by using a module function wrapper.]{\hspace{2.5em}\epsfig{file=diagrams/module_wrapper.eps,width=0.125\textwidth}} & %
\subfloat[\label{fig:module_probe}Module-only instrumentation, with a kernel function specialized to also be instrumented by using a probe-based wrapper.]{\hspace{2.5em}\epsfig{file=diagrams/module_probe.eps,width=0.125\textwidth}}
\end{tabularx}
\caption{Visual example of using different granularities of instrumentation to selectively instrument or not instrument specific code.}
\end{figure}

@@

\begin{figure*}[ht!]
\lstset{language=C, tabsize=2, stepnumber=1}
\begin{multicols}{2}
\begin{lstlisting}[basicstyle=\scriptsize\ttfamily]
struct device_driver {
  ...
  int  (*probe)(struct device *);
  int  (*remove)(struct device *);
  void (*shutdown)(struct device *);
  int  (*suspend)(struct device *, pm_message_t);
  int  (*resume)(struct device *);
  ...
  const struct dev_pm_ops *pm;
  ...
};
\end{lstlisting}
\columnbreak
\begin{lstlisting}[basicstyle=\scriptsize\ttfamily]
TYPE_WRAPPER(struct device_driver, {
  PRE_OUT {
    ABORT_IF_FUNCTION_IS_WRAPPED(arg.probe)
    WRAP_FUNCTION(arg.probe);
    WRAP_FUNCTION(arg.remove);
    ...
  }
  POST_OUT {
    POST_WRAP(arg.pm);
  }
})
\end{lstlisting}
\end{multicols}
\caption{The Linux device driver structure is shown on the left. The automatically generated type wrapper for this structure is shown on the right. In the wrapper code, \texttt{arg} is a reference to a \texttt{struct device\_driver} object passed as, or referenced by, an argument to a kernel or module wrapper. Code in the \texttt{PRE\_OUT} section is applied to arguments of the wrapped type before a kernel wrapper is invoked. Similarly, code in the \texttt{POST\_OUT} section is applied to arguments of the wrapped type after a kernel wrapper is invoked. \texttt{POST\_WRAP} invokes the type wrapper that is specific to the value to which it is applied (\texttt{arg.pm}). Type wrappers also support \texttt{\_IN} suffixes instead of \texttt{\_OUT} suffixes, which apply to data going into modules (i.e., over module wrappers). Finally, the \texttt{RETURN\_} prefix is used to apply some code to return values of either kernel or module wrappers.}
\label{fig:type_wrapper}
\end{figure*}

Unlike DRK and btkernel, Granary can target instrumentation only at specific kernel modules, while executing the rest of the kernel natively and without overhead. Instrumenting only specific kernel modules is particularly useful for debugging tools because modules represent the bulk of new kernel code under development, and contain the most bugs \cite{FaultsInLinux}.

Granary interposes on an existing module by using hardware page protection to mark an existing module as non-executable and trap attempts to execute that module's code. Granary transparently recovers from these traps by returning execution to instrumented code. Granary takes control of dynamically loaded modules by interposing on the Linux kernel's module loading process. When a kernel module is loaded, Granary bootstraps by translating the first basic block of the module's initialization function.  It then replaces the pointer to that function with a pointer to the translated basic block. The translation process continues when the kernel initializes the module by invoking the translated module code. As with existing modules, the native code segments of dynamically loaded modules are page protected to prevent and recover from accidental execution.

While correct, hardware page protection is an inefficient mechanism for gaining control of module code execution. That is, after the (instrumented) module initializer runs to completion, Granary needs an efficient way to regain control of future invocations of the module's code. Our approach to regaining control is based on the observation that modules tell the kernel about their interfaces by registering functions with the kernel. We expect that at least some of the registered functions will be executed by the kernel because this is the mechanism by which modules extend the kernel's functionality. To Granary, registered module functions represent potential future \emph{attach} points, where a switch from native to instrumented execution will occur. Furthermore, Granary discovers additional attach points at detach points by observing pointers to module functions that are passed by modules to the kernel.

Granary uses static analysis to dynamically discover attach points by wrapping the kernel/module interface. Kernel functions are wrapped by \emph{kernel wrappers} that inspect and traverse argument pointers in search of pointers to module functions. \Cref{fig:type_wrapper} shows an example of a type wrapper for the Linux device driver structure. If a pointer to a module function (a future attach point) is discovered, then Granary replaces that pointer with a function-specific \emph{module wrapper}. Granary inspects and modifies arguments to module wrappers in the same way as for kernel wrappers. This allows Granary to discover kernel entry points that will cause instrumentation to detach. Finally, Granary redirects execution to the appropriate kernel or instrumented module function after wrapping has occurred.

Kernel and module wrappers invoke \emph{type wrappers} to find and wrap pointers to module functions that are directly or indirectly referenced by kernel/module function arguments. A type wrapper is a function that recursively  traverses the in-memory object graph and converts pointers to module code into pointers to wrapped module functions. Type and kernel wrappers are automatically generated at Granary's compile time by scripts that statically analyze the kernel source code. Granary automatically matches any variable in a kernel wrapper to a type wrapper if the base type (absent pointers, specifiers, and qualifiers) of that variable matches the type wrapper's wrapped type. Similarly, Granary automatically generates module wrappers using a combination of compile-time meta-programming and runtime code generation to match type wrappers to the declared arguments of \texttt{C} function pointer types.  Granary's wrapper approach has two benefits: (i) it is more efficient than the trap-based alternative, and (ii) it gives instrumentation tools direct and semantically meaningful access to data crossing the kernel/module boundary.

\label{sec:implicit_attach}
After Granary bootstraps on the module initialization process, \emph{attaching} occurs in one of three ways: \begin{enumerate}
	\item {\bf Implicit attaching:} the kernel returns to instrumented module code in the code cache. This occurs when instrumented moduel code is interrupted, or when the module invokes a kernel function (detaching) and that function returns to the instrumented module code (attaching).
	\item {\bf Fast attaching:} the kernel invokes a wrapped module function. The first instance of fast attaching for any kernel module is when the kernel's module loader invokes the instrumented module initializer function. 
	\item {\bf Slow attaching:} the kernel invokes unwrapped module code. If the kernel executes a module function which was passed to the kernel in a type-unsafe manner, then the processor will raise a fault because Granary uses hardware page protection to prevent module code from being executed. Granary handles these faults by returning execution to the instrumented version of the faulting module code.
\end{enumerate}

\label{sec:detach}
Granary \emph{detaches} when control transfers from instrumented code to native (uninstrumented) code. Detaching occurs in one of two ways: \begin{enumerate}
	\item {\bf Implicit detaching:} instrumented module code returns to the kernel code which originally invoked the module, or is interrupted (initial interrupt handling is done by the kernel).
	\item {\bf Wrapped detaching:} instrumented module code invokes a kernel wrapper, which later transfers control to the kernel.
\end{enumerate}

>>>>

% Make text in figure captions small
\usepackage{caption3} % load caption package kernel first
\DeclareCaptionOption{parskip}[]{} % disable "parskip" caption option
\usepackage[small]{caption}

@@

% Compressed enumerations
\newenvironment{enumerate*}%
  {\begin{enumerate}%
    \setlength{\itemsep}{2pt}%
    \setlength{\parskip}{0pt}}%
  {\end{enumerate}}

@@

\section{Only Instrument What is Needed}\label{sec:what}

Some instrumentation tools have a limited scope, and therefore do not need to instrument all code. For example, Granary's Guardrail tool is concerned with finding bugs in Linux device drivers \cite{Guardrail}. Guardrail records traces of executed instructions, and analyzes those traces to find memory access and concurrency bugs. Both phases are heavyweight, and so limiting the scope of what is instrumented to what is of interest (module code) results in better performance. Unfortunately, existing systems are unable to change the granularity of instrumentation. Fortunately, Granary allows Guardrail to set the granularity of instrumentation to only instrument a specific module, while leaving the rest of the kernel to execute without overhead.

%The closest that an existing system can get to

% Therefore, if Guardrail were implemented using DRK or btkernel, then it would have to implement its own form of runtime code specialization

%. Recording and analyzing every instruction executed by the kernel is

%Guardrail records and analyzes traces of device driver code to find bugs. However, Guardrail's analysis can be heavyweight. If Guardrail were forced to instrument all code (e.g. with DRK or btkernel), then Guardrail would impose significant performance overheads. However, if Guardrail's recording were limited to only module code (

%and so limiting the analysis to only device driver code results in better performance.

@@

Guardrail is an example of how tools can choose the instrumentation granularity that best matches their needs. That is, a Granary tool can choose to instrument some code, while leaving other code to execute natively. This is useful when the instrumentation in question has a high overhead or when only a subset of the kernel's code is of interest. \Cref{tab:granularities} provides a non-exhaustive list of the different granularities of instrumentation supported by Granary.

@@

\subsection{Instrumenting All Kernel Code}

Like DRK and btkernel, Granary can instrument the whole kernel. Instrumenting the whole kernel requires interposing on the system call entrypoint\footnote{The system call entrypoint is stored in the x86-64 \texttt{MSR\_LSTAR} model-specific register.} and on the interrupt vector entrypoints\footnote{Interrupt vector entrypoints are stored in the interrupt descriptor table, which can be read from and written to using the \texttt{sidt} and \texttt{lidt} instructions, respectively.}. For whole-kernel instrumentation, Granary bootstraps by duplicating the the early entrypoint instructions that are responsible for switching execution onto a kernel stack, and then injects a jump to translated/instrumented kernel code after the stack switch. The translation process continues when a system call or interrupt occurs and follows the jump from duplicated into translated code. Granary only translates code that is known to operate on a kernel stack\footnote{Not all kernel code executes on a kernel stack. For example, when a user space application invokes the \texttt{syscall} instruction, control is transferred to the kernel, but no stack switch occurs. The kernel's early system call entrypoint code is responsible for switching execution onto a kernel stack.} so that instrumentation tools can safely save transient data in a reentrant way by spilling that data to the stack.

\begin{figure}[t]
\captionsetup[subfloat]{width=0.45\columnwidth}
\begin{tabularx}{\columnwidth}{XX}%
\subfloat[Full-kernel instrumentation.]{\hspace{2.5em}\epsfig{file=diagrams/kernel.eps,width=0.125\textwidth}} & %
\subfloat[Module-only instrumentation]{\hspace{2.5em}\epsfig{file=diagrams/module.eps,width=0.125\textwidth}} \\
\subfloat[Module-only instrumentation, with a module function wrapped to run natively.]{\hspace{2.5em}\epsfig{file=diagrams/module_wrapper.eps,width=0.125\textwidth}} & %
\subfloat[\label{fig:module_probe}Module-only instrumentation, with a kernel function wrapped to also be instrumented.]{\hspace{2.5em}\epsfig{file=diagrams/module_probe.eps,width=0.125\textwidth}}
\end{tabularx}
\caption{Visual example of using different granularities of instrumentation to selectively instrument or not instrument specific code.}
\end{figure}

@@

Granary is the first kernel-space instrumentation system able to target instrumentation only at specific kernel modules, while executing the rest of the kernel natively and without overhead. Instrumenting only specific kernel modules is particularly useful for debugging tools because modules represent the bulk of new kernel code under development, and contain the most bugs \cite{FaultsInLinux}.

The basic technique that Granary uses to interpose on module code is hardware page protection. First, Granary page protects existing module code, marking it as non-executable. Then, Granary traps attempts to execute the protected module code and transparently recovers from these traps by returning execution to instrumented code. Granary also interposes on the Linux kernel's module loading process. When a kernel module is dynamically loaded, Granary bootstraps by translating the first basic block of the module's initialization function.  It then replaces the pointer to that function with a pointer to the translated basic block. The translation process continues when the kernel initializes the module by invoking the translated module code. As with existing modules, the native code segments of dynamically loaded modules are page protected to prevent and recover from accidental execution.

%Granary interposes on an existing module by using hardware page protection to mark the module's code as non-executable and trapping attempts to execute the protected code.

While correct, hardware page protection is an inefficient mechanism for gaining control of module code execution. That is, after the (instrumented) module initializer runs to completion, Granary needs an efficient way to regain control of future invocations of the module's code. Our approach to regaining control is based on the observation that modules tell the kernel about their interfaces by registering interface functions with the kernel. These functions are the only supported mechanism for extending the kernel's functionality in a standardized way. When the kernel executes registered functions, control transfers from the kernel into the module. To Granary, registered module functions represent potential future \emph{attach} points, where a switch from native to instrumented execution will occur. Furthermore, Granary discovers additional attach points at detach points by observing pointers to module functions that are passed by modules to the kernel.

Granary uses static analysis to dynamically discover potential attach points. Potential attach points include: \begin{enumerate*}
	\item Function pointers passed as arguments to kernel functions.
	\item Function pointers stored in structures passed to kernel functions. This rule applies transitively. For example, if one structure has a field that points to another structure, and the latter structure contains a function pointer, then the latter structure's function pointers are also potential attach points. This is a common pattern in the Linux kernel, and mimics virtual tables in object-oriented languages.
	\item Function pointers returned from attach points.
\end{enumerate*}

%Any function pointer passed as an argument to a kernel function, or stored in a structure that can be passed as an argument to a kernel function is treated as a potential attach point.

The static analysis phase of Granary generates ``wrappers'' for the kernel/module interface. Interface wrappers come in two forms: kernel function wrappers and data-structure wrappers (called type wrappers). Kernel function wrappers are transparently invoked when module code attempts to invoke kernel code. Function wrappers delegate to type wrappers to find attach points passed to the kernel.

%\begin{figure*}[t!]
%\lstset{language=C, tabsize=2, stepnumber=1}
%\begin{multicols}{2}
%\begin{lstlisting}[basicstyle=\scriptsize\ttfamily]
%struct device_driver {
%  ...
%  int  (*probe)(struct device *);
%  int  (*remove)(struct device *);
%  void (*shutdown)(struct device *);
%  int  (*suspend)(struct device *, pm_message_t);
%  int  (*resume)(struct device *);
%  ...
%  const struct dev_pm_ops *pm;
%  ...
%};
%\end{lstlisting}
%\columnbreak
%\begin{lstlisting}[basicstyle=\scriptsize\ttfamily]
%TYPE_WRAPPER(struct device_driver, {
%  PRE_OUT {
%    ABORT_IF_FUNCTION_IS_WRAPPED(arg.probe)
%    WRAP_FUNCTION(arg.probe);
%    WRAP_FUNCTION(arg.remove);
%    ...
%  }
%  POST_OUT {
%    POST_WRAP(arg.pm);
%  }
%})
%\end{lstlisting}
%\end{multicols}
%\caption{The Linux device driver structure is shown on the left. The automatically generated type wrapper for this structure is shown on the right. In the wrapper code, \texttt{arg} is a reference to a \texttt{struct device\_driver} object passed as, or referenced by, an argument to a kernel or module wrapper. Code in the \texttt{PRE\_OUT} section is applied to arguments of the wrapped type before a kernel wrapper is invoked. Similarly, code in the \texttt{POST\_OUT} section is applied to arguments of the wrapped type after a kernel wrapper is invoked. \texttt{POST\_WRAP} invokes the type wrapper that is specific to the value to which it is applied (\texttt{arg.pm}). Type wrappers also support \texttt{\_IN} suffixes instead of \texttt{\_OUT} suffixes, which apply to data going into modules (i.e., over module wrappers). Finally, the \texttt{RETURN\_} prefix is used to apply some code to return values of either kernel or module wrappers.}
%\label{fig:type_wrapper}
%\end{figure*}

\begin{figure}[t]
\lstset{language=C, tabsize=2, stepnumber=1}
\begin{lstlisting}[basicstyle=\scriptsize\ttfamily]
// struct inode *
// iget_locked(struct super_block *, unsigned long);
FUNCTION_WRAPPER(MODULE, iget_locked, (sb, ino), {
  PRE_OUT_WRAP(sb);
  struct inode *inode = iget_locked(sb, ino);
  POST_OUT_WRAP(sb);
  RETURN_IN_WRAP(inode);
  return inode;
})
\end{lstlisting}
\caption{Atomatically generated wrapper for the \texttt{iget\_locked} kernel function. \texttt{iget\_locked} obtains an inode from a mounted file system. The super block argument (\texttt{sb}) is first pre-wrapped before it is passed to the original \texttt{iget\_locked} function. After \texttt{iget\_locked} returns, \texttt{sb} is then post-wrapped. Finally, the inode data structure returned by the kernel (\texttt{inode}) is return-wrapped and then returned to instrumented module code. The \texttt{OUT} and \texttt{IN} qualifiers of the wrappers denote the direction that data is travelling. \texttt{OUT}-wrappers apply to data passed from instrumented code out to the native code. \texttt{IN}-wrappers apply to data passed from native code into instrumented code.}
\end{figure}

%\Cref{fig:type_wrapper} shows an example of a type wrapper for the Linux device driver structure.

If a pointer to a module function (a future attach point) is discovered, then Granary replaces that pointer with a function-specific \emph{module function wrapper}. Module function wrappers behave similarly to kernel function wrappers, in that they provide instrumentation tools a mechanism to observe a

%are responsible for wrapping their arguments. When an argument

%At runtime, these wrappers interpose on kernel functions and traverse in-memory data structures, find attach points, and convert them into

%At runtime, Granary detects and interposes on potential attach points by wrapping the kernel/module interface.

%Kernel functions are wrapped by \emph{kernel wrappers} that inspect and traverse argument pointers in search of pointers to module functions.  Granary inspects and modifies arguments to module wrappers in the same way as for kernel wrappers. This allows Granary to discover kernel entry points that will cause instrumentation to detach. Finally, Granary redirects execution to the appropriate kernel or instrumented module function after wrapping has occurred.

%Kernel and module wrappers invoke \emph{type wrappers} to find and wrap pointers to module functions that are directly or indirectly referenced by kernel/module function arguments. A type wrapper is a function that recursively  traverses the in-memory object graph and converts pointers to module code into pointers to wrapped module functions. Type and kernel wrappers are automatically generated at Granary's compile time by scripts that statically analyze the kernel source code. Granary automatically matches any variable in a kernel wrapper to a type wrapper if the base type (absent pointers, specifiers, and qualifiers) of that variable matches the type wrapper's wrapped type. Similarly, Granary automatically generates module wrappers using a combination of compile-time meta-programming and runtime code generation to match type wrappers to the declared arguments of \texttt{C} function pointer types.  Granary's wrapper approach has two benefits: (i) it is more efficient than the trap-based alternative, and (ii) it gives instrumentation tools direct and semantically meaningful access to data crossing the kernel/module boundary.

%\label{sec:implicit_attach}
%After Granary bootstraps on the module initialization process, \emph{attaching} occurs in one of three ways: \begin{enumerate}
%	\item {\bf Implicit attaching:} the kernel returns to instrumented module code in the code cache. This occurs when instrumented moduel code is interrupted, or when the module invokes a kernel function (detaching) and that function returns to the instrumented module code (attaching).
%	\item {\bf Fast attaching:} the kernel invokes a wrapped module function. The first instance of fast attaching for any kernel module is when the kernel's module loader invokes the instrumented module initializer function. 
%	\item {\bf Slow attaching:} the kernel invokes unwrapped module code. If the kernel executes a module function which was passed to the kernel in a type-unsafe manner, then the processor will raise a fault because Granary uses hardware page protection to prevent module code from being executed. Granary handles these faults by returning execution to the instrumented version of the faulting module code.
%\end{enumerate}
%
%\label{sec:detach}
%Granary \emph{detaches} when control transfers from instrumented code to native (uninstrumented) code. Detaching occurs in one of two ways: \begin{enumerate}
%	\item {\bf Implicit detaching:} instrumented module code returns to the kernel code which originally invoked the module, or is interrupted (initial interrupt handling is done by the kernel).
%	\item {\bf Wrapped detaching:} instrumented module code invokes a kernel wrapper, which later transfers control to the kernel.
%\end{enumerate}
